---
title: Changes in the Canine Plasma Lipidome after Short- and Long-Term Excess Glucocorticoid Exposure 
subtitle: Data Analysis for the Manuscript 'Sieber-Ruckstuhl, Burla, et al., 2019'
author: 
  - name: Nadja S. Sieber-Ruckstuhl, Bo Burla, Susanne Spoerel, Florence Schmid, Claudio Venzin, Amaury Cazenave-Gassiot, Anne K. Bendt, Federico Torta, Markus R. Wenk, Felicitas S. Boretti
    affiliation: Vetsuisse Faculty, University of Zurich and Singapore Lipidomics Incubator (SLING), National University of Singapore
date: February 25, 2019 (B.Burla)
output:
  html_document:
    toc: yes
  pdf_document: 
    latex_engine: lualatex
  word_document: default
---

```{r Init, message=FALSE, warning=FALSE, include=TRUE}
###################################################################################################
# Data processing, statistics and plots for the Manuscript 'Sieber-Ruckstuhl, Burla, et al., 2019'
#
# 20.01.2019 / Bo Burla 
# Singapore Lipidomics Incubator (SLING), National University of Singapore
###################################################################################################

library(here)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(tibble)
library(gplots)
library(RColorBrewer)
library(ggpubr)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(factoextra)
library(ComplexHeatmap)   # Required version V1.99 or higher
library(circlize)
library(knitr)
library(ggsignif)
library(readr)

source(here("analysis/ms_data_preprocessing.R"))
source(here("analysis/misc_functions.R"))

```
## 1) Figures 1 a-l: Clinical plasma markers of lipid metabolism
```{r Fig1, echo=TRUE, fig.align='center', fig.height=10, fig.width = 4.5, message=FALSE, warning=TRUE}
data_file <- here("data","raw","Sieber-Ruckstuhl_Burla_2019_RawData-Annot.xlsx") 
d_clinchem_wide <- read_excel(data_file, sheet = "ClinicalChemistry",
                              col_names = TRUE,trim_ws = TRUE,na = c("#N/A", "NULL"), col_types = "guess")

#Conver to long format, use only parameter with numerical value 
d_clinchem <- d_clinchem_wide %>% 
  select(AnimalID, TreatmentGroup, TreatmentTimePoint, Cortisol, Lipase, TG, Cholesterol, ALAT, ALP) %>% 
  gather(key = "Parameter", value =  "Value", -AnimalID:-TreatmentTimePoint, factor_key=TRUE,convert = TRUE)

# re-order the levels in the order of appearance in the dataframe otherwise it plots it in alphabetical order
d_clinchem$TreatmentTimePoint <- factor(d_clinchem$TreatmentTimePoint, levels=c("before","after"))
d_clinchem$AnimalID <- as.factor(d_clinchem$AnimalID)
d_clinchem$TreatmentGroup <- as.factor(d_clinchem$TreatmentGroup)
d_clinchem$Parameter <- as.factor(d_clinchem$Parameter)
d_clinchem$Value <- as.numeric(d_clinchem$Value)

# Create labels to label points 
d_clinchem <- d_clinchem %>% mutate(label=ifelse(TreatmentTimePoint=="before", "", as.character(AnimalID)))  
t.test.log <- function(x,y){t.test(x=log2(x*10),y=log2(y*10),paired = TRUE,var.equal = FALSE)}
  
paired_plot <- function(d_filt, var,plotY){
  d_filt_param <- d_filt %>% filter(Parameter==var)
  p<- ggplot(d_filt_param, aes(x=TreatmentTimePoint, y=Value, group=AnimalID)) +
      geom_point(aes(colour=TreatmentTimePoint), size=1.5) +
      geom_line(size=0.5, alpha=0.5,  linetype="longdash") +
      xlab('') +
      scale_colour_manual(values=c("black", "black"), guide=FALSE) +
      ylab(ifelse(plotY, var, "")) +
      scale_y_continuous(expand = c(0.1,0.0)) +
      scale_x_discrete(expand = c(0.1,0.2)) +
      aes(ymin=0) +
      geom_text_repel(label=as.character(d_filt_param$label), size=2, nudge_x=.1, nudge_y=0.1) +
      geom_signif(comparisons  = list(c("before", "after")), 
                  textsize = 2, test = "t.test.log", map_signif_level = FALSE,margin_top = 0.15) +
      theme_pubr() +
      theme(axis.title = element_text(face = "bold"), axis.text.y = element_text(size=10)) +
      theme(legend.position="none") +
      theme(strip.background = element_blank(),strip.text.x = element_blank())
  return(p)
}


# Create list of plots of parameters from one treament group, remove x-axis text from all except last plot
plt_PRED <- map(.x = c("Cortisol", "Lipase", "TG", "Cholesterol", "ALAT", "ALP"),  
                .f = function(x) paired_plot(d_clinchem[d_clinchem$TreatmentGroup=="PRED",],x, TRUE))
plt_PRED[-length(plt_PRED)] <- map(plt_PRED[-length(plt_PRED)], function(x) x + rremove("x.text"))


plt_ACTH <- map(.x = c("Cortisol", "Lipase", "TG", "Cholesterol", "ALAT", "ALP"),  
                .f =function(x) paired_plot(d_clinchem[d_clinchem$TreatmentGroup=="ACTH",],x, FALSE))
plt_ACTH[-length(plt_ACTH)] <- map(plt_ACTH[-length(plt_ACTH)], function(x) x + rremove("x.text"))

# Create a grid from all plots
fig1 <- ggarrange(plotlist=c(rbind(plt_PRED,plt_ACTH)), ncol = 2, nrow = 6, align="h")
ggsave(filename = here("output/figure_1_clinchem.pdf") ,plot = fig1, width = 4, height = 10,units = "in")

plot_grid(plotlist = c(rbind(plt_PRED,plt_ACTH)), align='vh',ncol = 2)
```

## 2) Load and process MS-based lipidomics data 
Load raw peak area data, normalize with internal standards, calculate concentrations and perform QC filtering)
```{r LoadData, echo=TRUE, fig.height=10, fig.width=5, fig.align='center', message=FALSE, warning=TRUE}
#d_conc_wide <- read_csv(here("data/processed/Ruckstuhl_Burla_2019_BeaglePlasmaLipididomeAfterCS_LipidConc_uM.csv"),  col_names = TRUE, trim_ws =TRUE)

data_annot_file = here("data/raw/Sieber-Ruckstuhl_Burla_2019_RawData-Annot.xlsx") 
d_conc_wide <- process_ms_dataset(data_annot_file) %>% 
                arrange((TreatmentGroup),desc(TreatmentTimePoint),  AnimalID)   

write_csv(d_conc_wide, 
            here("output", "table_S2_LipidConc_uM_Ruckstuhl_Burla_2019.csv"),
            col_names = TRUE)

d_conc <- d_conc_wide %>% gather(key = Compound, value = Conc, -SampleName:-Sex, factor_key=TRUE) %>%
          add_lipidclass_names() 
```

## 3) Figures 2 a/b: Heatmaps and clustering analyses of lipidomes
```{r Fig2AB, echo=TRUE, fig.show='hold', fig.align='center', fig.height=7, fig.width=5, message=TRUE, warning=TRUE}
heatmap_plot <- function(d, grp_colors){

    # Prepare data matrix: Remove annotations remove lipid species containing NAs, Inf and zeros
  d_conc_only <- d %>% select(-SampleName:-Sex) %>% 
                    select_if(~ !any(is.na(.)) & !any(is.infinite(.)) & !any(.==0)) 
  
  # Log-transform, scale and center data
  m_conc_log <- as.matrix(log2(d_conc_only*100000))
  m_conc_log_norm = scale(m_conc_log, center = TRUE, scale = TRUE)
  rownames(m_conc_log_norm) <- paste0(d$AnimalID,d$TreatmentTimePoint)
  
  m_conc_log_norm_t <- t(m_conc_log_norm)
  
  # cluster samples and compoundss
  d1 <- dist(m_conc_log_norm,method = "manhattan", diag = TRUE, upper = TRUE)
  d2 <- dist(m_conc_log_norm_t,method = "maximum", diag = TRUE, upper = FALSE)
  c1 <- hclust(d1, method = "ward.D", members = NULL)
  c2 <- hclust(d2, method = "ward.D2", members = NULL)
  
  # Create column with color annotations
  color.map <- ifelse(d$TreatmentTimePoint=="after",grp_colors[1], grp_colors[2])
  
  breaks <- seq(-3,3,length=256)
  # Color gradient
  my_palette <- rev(colorRampPalette(brewer.pal(10, "RdBu"))(255))
  
  # plot heatmap
  heatmap.2(m_conc_log_norm_t,trace="none",
        Colv=as.dendrogram(c1),     
        Rowv=as.dendrogram(c2),
        scale="row",
        col = my_palette,
        ColSideColors=color.map, breaks=breaks,
        dendrogram = "both", xlab = NULL, ylab = NULL, 
        margins=c(8,7),cexRow=0.4, cexCol = 1.4, density.info="none",labRow = FALSE, 
        key = TRUE)
}

pdf(file = here("output/figure_2ab_clustering.pdf"), width = 5.5, height = 7)
  heatmap_plot(d_conc_wide %>% filter(TreatmentGroup == "ACTH"), c("#E22D01","#ffeeea"))
  heatmap_plot(d_conc_wide %>% filter(TreatmentGroup == "PRED"), c("#0F85DC","#c9dfef"))
dev.off()

heatmap_plot(d_conc_wide %>% filter(TreatmentGroup == "ACTH"), c("#E22D01","#ffeeea"))
heatmap_plot(d_conc_wide %>% filter(TreatmentGroup == "PRED"), c("#0F85DC","#c9dfef"))
```

## 4) Figures 2 c/d:  Principal Component Analyses (PCA) of lipidomes
```{r Fig2CD, echo=TRUE, fig.height=5, fig.width=6, fig.align='center', message=FALSE, warning=TRUE}
d_conc_pcaplot <- d_conc_wide %>% select(-SampleName:-Sex) %>% 
                          select_if(~ !any(is.na(.)) & !any(is.infinite(.)) & !any(.==0)) 

# Log2 transform the data
d_conc_log_pcaplot <- log2(d_conc_pcaplot*100000)

# Perform PCA and plot the results
res_pca <- prcomp(d_conc_log_pcaplot, center = TRUE, scale. = TRUE)

pca_plot <- function(comp_x, comp_y){
  fviz_pca_ind(res_pca, 
    axes = c(comp_x,comp_y),
    palette=c("#E22D01","#E22D01", "#0F85DC", "#0F85DC"),  
    habillage=d_conc_wide$ExperimentalGroup,
    addEllipses=TRUE, 
    ellipse.level=0.8,
    ellipse.alpha = 0,
    ellipse.line.size=1,
    mean.point = FALSE,
    pointsize = 6.2,
    label="none",
    title = "",
    axes.linetype = "dotted") +
  scale_shape_manual(values = c(0, 15, 2,17)) +
  geom_text_repel(aes(label=as.character(d_conc_wide$AnimalID)), size=4, nudge_x=0.6, nudge_y=0.6)+
  mytheme
}
fig_1C <- pca_plot(comp_x = 1,comp_y = 2)           
ggsave(filename = here("output/figure_2c_PCA-dim12.pdf"),plot = fig_1C, width = 6.2, height = 5)
fig_1C

fig_1D <- pca_plot(comp_x = 1, comp_y = 3)
ggsave(filename = here("output/figure_2d_PCA-dim13.pdf"),plot = fig_1D, width = 6.2, height = 5)
fig_1D 
```

## 5) Calculate fold-changes and P values before/after treatment
Calculate individual and average fold changes before/after treatments, and FDR-adjusted P values from paired t-tests
```{r ProcessData, echo=TRUE, fig.height=10, fig.width=5, fig.align='center', message=FALSE, warning=TRUE}
get_FC_stat_summary <- function(d_Conc_local){
  # Get data frame with fold changes before vs. after treatment for each species/experimental group
  d_FC <- d_Conc_local %>% 
  group_by(TreatmentGroup, AnimalID, Sex, lipidClassBase, lipidClass, isPUFA, Compound, add = FALSE)  %>% 
    summarise(log2FC = log2(Conc[TreatmentTimePoint=="after"]/Conc[TreatmentTimePoint=="before"])) 

  # Obtain P value from paired t-test (incomplete cases are removed, i.e. species with one or more zero or NA)
  d_FC_stat <- d_Conc_local %>%
    group_by(Compound, TreatmentGroup) %>% 
       mutate (log2Conc = log2(Conc)) %>% 
       filter(sum(is.infinite(log2Conc) | is.na(log2Conc)) == 0) %>%   
       do(summarise(.,p_value_tLOG = t.test(log2Conc ~ TreatmentTimePoint,
                                            var.equal=FALSE, paired=TRUE)$p.value))
  # Calculate FDR-adjusted P values
  d_FC_stat <- d_FC_stat %>%
    group_by(TreatmentGroup, add = FALSE) %>%
     do(mutate(.,fdr_tLOG = p.adjust(p_value_tLOG, "fdr")))
  
  # Calculare average fold changes and combine with P values
  d_FC_summary <- d_FC %>% 
    group_by(Compound, lipidClass, lipidClassBase, TreatmentGroup) %>%
      mutate(log2FC = ifelse(is.infinite(log2FC),NA,log2FC)) %>%
       summarise(meanLog2FC = mean(log2FC, na.rm=TRUE),
                 sdLog2FC = sd(log2FC, na.rm=TRUE),
                 meanFC = 2^abs(meanLog2FC)*sign(meanLog2FC)) %>% 
    left_join(d_FC_stat)
  return(list("FC" = d_FC, "summary" = d_FC_summary))
}

# Calculate comparisons for the complete dataset (used in Figures 2 e/f and Figures 3-6)
fc_summary_all <- get_FC_stat_summary(d_conc)

```

## 6) Figures 2 e/f: PCA of lipid fold changes in all indiviuals
```{r Fig2EF, echo=TRUE, fig.height=5, fig.width=6, fig.align='center', message=FALSE, warning=TRUE}
d_FC_wide  <- fc_summary_all$FC %>% ungroup() %>% 
                select(-lipidClassBase, -lipidClass,-isPUFA) %>% 
                spread(key = Compound , value = log2FC) %>% 
                select_if(~ !any(is.na(.)) & !any(is.infinite(.)) & !any(.==0)) 

# Perform PCA and plot the results
res_pca <- prcomp(d_FC_wide %>% select(-TreatmentGroup:-Sex), center = TRUE, scale. = TRUE)


d_FC_wide$TreatmentGroup_shape <- ifelse(d_FC_wide$TreatmentGroup == "ACTH", 15, 17)

pca_plot_FC <- function(comp_x, comp_y, habillage, group_colors){
              fviz_pca_ind(res_pca, 
                axes = c(comp_x,comp_y),
                palette=group_colors,   
                habillage=factor(as.factor(habillage)),
                addEllipses=TRUE, 
                ellipse.level=0.8,
                ellipse.alpha = 0,
                ellipse.line.size=1,
                mean.point = FALSE,
                pointsize = 6,
                pointshape = d_FC_wide$TreatmentGroup_shape,
                pointcolour = d_FC_wide$TreatmentGroup,
                label="none",
                title = "",
                legend = "Groups",
                axes.linetype = "dotted") +
              scale_shape_manual(values = c(15, 17)) +
              geom_text_repel(aes(label=as.character(d_FC_wide$AnimalID)), size=4, nudge_x=0.6, nudge_y=0.6)+
              mytheme
            }
fig_1E <- pca_plot_FC(comp_x = 1, comp_y = 2, 
                      habillage = d_FC_wide$TreatmentGroup, 
                      group_colors = c("#E22D01","#0F85DC"))

fig_1F <- pca_plot_FC(comp_x = 3, comp_y = 4,
                      habillage = d_FC_wide$Sex, 
                      group_colors = c("#923D97","#0CB1A4"))  

ggsave(filename = here("output/figure_2e_PCA_FC_dim12.pdf"),plot = fig_1E, width = 5.88, height = 5)
fig_1E

ggsave(filename = here("output/figure_2f_PCA_FC_dim34.pdf"),plot = fig_1F, width = 5.65, height = 5)
fig_1F
  
```
Warnings are from fviz_pca_ind, with no effect on the data, figure (arises from line 'pointshape = d_FC_wide$TreatmentGroup_shape'). Fill color of points from individuals were changed in Illustrator to colors of experimental groups (see Fig. 2E) 

## 7) Figure 5 b: Total levels of phospholipid species with ≤3 and ≥4 double bonds
P values from t-tests comparing total levels of PUFA and non-PUFA containg species before and after treatments are shown after the figure.
```{r Fig5b, echo=TRUE, fig.height=5, fig.width=8, message=FALSE, warning=TRUE}
# Calculate total levels per lipid class per individual
# To be able to use function "get_FC_stat_summary" variables Conc/Compound are assigned with the total conc/lipid class


plot_nonPUFA_vs_PUFA <- function(d_conc, filename ){
  d_conc_lipidclass <- d_conc %>% 
  group_by(TreatmentGroup, AnimalID, Sex, lipidClass, isPUFA, TreatmentTimePoint) %>%
  summarise(Conc = sum(Conc,na.rm = TRUE), 
            Compound = paste0(lipidClass[1],ifelse(isPUFA[1], "_PUFA","_noPUFA" )), 
            lipidClassBase = lipidClassBase[1]) %>% ungroup()

  fc_summary_lipidClasses <- get_FC_stat_summary(d_conc_lipidclass)
  d_FC_temp <- fc_summary_lipidClasses$FC %>% 
    filter(lipidClass == "PC" | lipidClass == "PE"| lipidClass == "PI", TreatmentGroup == "ACTH")
  
  my_comparisons <- list( c("FALSE.PC", "TRUE.PC"), c("FALSE.PE", "TRUE.PE"), c("FALSE.PI", "TRUE.PI"))
  
  fig_5B <- ggplot(d_FC_temp, aes( x=interaction(isPUFA, lipidClass), y = log2FC )) +
    geom_point(size=3.3,alpha=0.7, aes(shape=isPUFA), color="#142EAF", fill="#142EAF") +
    geom_line(size=1, aes(group = interaction(AnimalID, lipidClass)),linetype="dotted", color="#BCCDCE") +
    scale_shape_manual(values=c(16,16)) +
    scale_fill_manual(values=c("#142EAF", "#142EAF")) +
    scale_color_manual(values=c("#142EAF", "#142EAF")) +
    stat_summary(fun.data="plot.mean", geom="errorbar", width=0.5, size=.3, color="red", 
                 inherit.aes=TRUE,position=position_dodge(0.6)) +
    geom_hline(yintercept = 0, linetype="dotted", size=1, color = "#919191") + 
    stat_compare_means(method = "t.test", paired = TRUE, comparisons = my_comparisons, label = "p.format") +
    xlab("")
  plot(fig_5B)

  
  pvalues <- fc_summary_lipidClasses$summary %>% ungroup() %>% 
    select(-lipidClassBase, -fdr_tLOG) %>% 
    filter(lipidClass == "PC" | lipidClass == "PE"| lipidClass == "PI", TreatmentGroup == "ACTH")
  ggsave(filename = here("output", filename), plot = fig_5B, width = 6, height = 3.7)
  knitr::kable(pvalues)
}

plot_nonPUFA_vs_PUFA(d_conc, "figure_5b_nonPUFAvsPUFA.pdf")

```


## 8) Figures 3/4/5/6: Heatmaps with individual/average fold changes and P / FDR values 
```{r Fig3456, echo=TRUE, fig.height=7.8, fig.width=4.4, message=FALSE, warning=TRUE, paged.print=FALSE}
# Each final heatmap consists of 3 fused heatmaps for meanFC, P value and individual FC

# Define color palette and legend for FC 
FC_palette <- colorRamp2(breaks = c(-2,-1, 0, 1, 2), colors = rev(brewer.pal(5, "RdBu")),space = "RGB")
fc_lgd = Legend(title = "log2(FC)", col = FC_palette, at = c(-2,-1,0,1,2), 
                labels = c("-2", "-1", "0" , "1", "2"),legend_height= unit(5, "cm"))

# Define color palette and legend for P values
pvalue_palette = colorRamp2(breaks = c(0,-log10(0.2), 1,-log10(0.05), 2,-log10(0.002)), 
                              colors = c("#ededed", "#ededed", "#f2fff3", "#7ff48c", "#07E012", "#017002"),
                              space = "RGB") 
pvalue_lgd = Legend(title = "p-value", col = pvalue_palette, 
                      at = c(-log10(0.2), 1,-log10(0.05), 2,-log10(0.002)), 
                      labels = c("0.2", "0.1", "0.05", "0.01", "0.002"),legend_height= unit(5, "cm"))

# Function generating the complete heatmap with columns for average FC and P value
plot_heatmap <- function(fc_summary_object, cmpd_filter, treatment_group_filter, show_row_names){
  
  # Constants
  LOG_FC_MAX_PLOT = 2
  
  # Used to set name of ComplexHeatmaps (they cannot contain numbers)
  cmpd_filter_charonly =  str_extract(cmpd_filter, "([a-zA-Z])+")
  
  # Filter dataset for specified lipd class and group and convert to matrix
  m_IndFC <- fc_summary_object$FC %>% mutate(log2FC = ifelse(is.infinite(log2FC),NA,log2FC)) %>% 
                        filter(TreatmentGroup==treatment_group_filter, str_detect(Compound, cmpd_filter)) %>% 
                        ungroup() %>%
                        select(-Sex, -TreatmentGroup, -lipidClass, -lipidClassBase, -isPUFA) %>% 
                        spread(key = AnimalID, value = log2FC) %>%
                        column_to_rownames("Compound") %>%
                        as.matrix() %>% t()
  class(m_IndFC) <- "numeric"
  
  # Obtain heatmap with individual fold changes 
  hm_IndvFC = Heatmap(t(m_IndFC), name = paste0("hm_IndvFC", cmpd_filter_charonly, treatment_group_filter),
              col = FC_palette,
              width = unit(nrow(m_IndFC) * 2.8 , "mm"),
              height = unit(ncol(m_IndFC) * 2.6, "mm"),
              show_row_names = FALSE,
              show_column_names = TRUE,
              column_names_rot = 90,
              column_names_gp = gpar(fontsize = 07, fontface="plain"),
              cluster_rows = FALSE,cluster_columns = FALSE,
              show_heatmap_legend = FALSE,
              rect_gp = gpar(col = "white", lty = 1, lwd = 0.3))
  
  # Prepare FC and P value matrix (i.e. truncate FC at +/- 2)  
  d_sum <- fc_summary_object$summary %>% 
    mutate(log_p_value = -log10(fdr_tLOG),
           fdr_tLOG = ifelse(is.na(fdr_tLOG), 1, fdr_tLOG),
           signif_text = cut(fdr_tLOG, c(0,0.001,0.01, 0.05, 0.1,1), 
                             labels=c("***","**","*","",""), right=TRUE),
           FC_text = as.character(ifelse(!is.nan(meanFC), sprintf("%0.1f", round(meanFC, digits = 1)),"")),
           meanlog2FC_cut = ifelse(abs(meanLog2FC)>LOG_FC_MAX_PLOT, 
                                   sign(meanLog2FC)*LOG_FC_MAX_PLOT, meanLog2FC)) %>%
    filter(TreatmentGroup==treatment_group_filter, str_detect(Compound, cmpd_filter)) 
  
  # obtain heatmap column with P values
  m_sum_p <- d_sum  %>% ungroup() %>% select(log_p_value) %>% rename( P = log_p_value)  %>%  as.matrix()
  rownames(m_sum_p) <- d_sum$Compound
  hm_pvalue= Heatmap(m_sum_p, name = paste0("hm_pvalue", cmpd_filter_charonly, treatment_group_filter),
                width = unit(5, "mm"),
                height = unit(ncol(m_IndFC) * 2.6, "mm"),
                col=pvalue_palette,  cluster_rows = FALSE,cluster_columns = FALSE,
                rect_gp = gpar(col = "white", lty = 1, lwd = 0.3),
                show_row_names = show_row_names,
                row_names_side = "left",
                row_names_gp = gpar(fontsize = 7),
                show_column_names = TRUE,
                column_names_rot = 90,
                column_names_gp = gpar(fontsize = 07, fontface="bold.italic"),
                show_heatmap_legend = FALSE,
                cell_fun = function(j, i, x, y, width, height, fill) {
                  grid.text(sprintf("%s", d_sum$signif_text[i]), x = x, y = y- unit(c(0.5), "mm"), 
                            gp = gpar(fontsize = 10))})
  
  # obtain heatmap column with average fold changes
  m_sum_FC <- d_sum  %>% ungroup()%>%  select(meanlog2FC_cut) %>% rowMeans() %>%  as.matrix() 
  colnames(m_sum_FC) <- c("FC")
  hm_meanFC = Heatmap(m_sum_FC, name = paste0("hm_meanFC", cmpd_filter_charonly, treatment_group_filter),
                col = FC_palette,
                width = unit(5, "mm"),
                height = unit(ncol(m_IndFC) * 2.6, "mm"),
                cluster_rows = FALSE,cluster_columns = FALSE,
                rect_gp = gpar(col = "white", lty = 1, lwd = 0.3),
                show_row_names = FALSE,
                show_column_names = TRUE,
                column_names_rot = 90,
                column_names_gp = gpar(fontsize = 07, fontface="bold"),
                show_heatmap_legend = FALSE,
                cell_fun = function(j, i, x, y, width, height, fill) {
                  grid.text(sprintf("%s", d_sum$FC_text[i]), x, y, gp = gpar(fontsize = 5))})
  
  # combine heatmaps 
  a <- draw(hm_pvalue + hm_meanFC + hm_IndvFC, gap = unit(c(0, 0.02), "cm"),  
            show_heatmap_legend = FALSE,
            newpage = FALSE)
  
  # draw borders around heatmap segments
  add_borders <- function(heatmap_name, borders_x, borders_y, line_width){
    decorate_heatmap_body(paste0(heatmap_name, cmpd_filter_charonly, treatment_group_filter), 
                          {grid.lines(borders_x, borders_y, gp = gpar(col = "#787878", lwd = line_width))})
  }
  
  add_borders("hm_IndvFC", c(0, 0), c(0,1),1.2)
  add_borders("hm_IndvFC", c(1, 1), c(0,1),1.2)
  add_borders("hm_IndvFC", c(0, 1), c(1,1),1.2)
  add_borders("hm_IndvFC", c(0, 1), c(0,0),1.2)
  add_borders("hm_pvalue", c(0, 1), c(1,1),1.2)
  add_borders("hm_pvalue", c(0, 1), c(0,0),1.2)
  add_borders("hm_pvalue", c(0, 0), c(0,1),1.2)
  add_borders("hm_meanFC", c(0, 1), c(1,1),1.2)
  add_borders("hm_meanFC", c(0, 1), c(0,0),1.2)
  add_borders("hm_meanFC", c(1, 1), c(0,1),1.2)
  draw(a)
  return(a)
}


# combine heatmaps for both experimental groups
plot_FC_heatmap_figure <- function(fc_summary_object, lipidclasses){
  i = 1
  plt_acth <- list()
  plt_pred <- list()
  for(lipidclass in lipidclasses){
    plt_pred[[i]] <- grid.grabExpr(plot_heatmap(fc_summary_object, lipidclass,"PRED", show_row_names = TRUE))
    plt_acth[[i]] <- grid.grabExpr(plot_heatmap(fc_summary_object, lipidclass,"ACTH", show_row_names = FALSE))
    i = i + 1
  }
  plot_grid(plotlist = c(rbind(plt_pred,plt_acth)), align='vh',axis = "r",ncol = 2)
}
 
# draw heatmaps for different lipid classes
pdf(file = here("output/figure_3-6_heatmaps.pdf"), width = 5 , height = 9)
  plot_FC_heatmap_figure(fc_summary_all, c("^Cer","Hex1Cer", "Hex2Cer", "^GM3"))
  plot_FC_heatmap_figure(fc_summary_all, c("SM", "S1P"))
  plot_FC_heatmap_figure(fc_summary_all, c("^LPC O","^PC O", "PC P", "^PE O","^PE P"))
  plot_FC_heatmap_figure(fc_summary_all, c("^LPC [0-9]+","^PC [0-9]+"))
  plot_FC_heatmap_figure(fc_summary_all, c("^LPE [0-9]+","^PE [0-9]+","^PI", "^PS"))
  plot_FC_heatmap_figure(fc_summary_all, c("DG ","TG ", "CE "))
  plot.new()
  draw(pvalue_lgd)
  plot.new()
  draw(fc_lgd)
dev.off()
  
```

## 9) Supplementary Table S3: Fold changes and p/FDR values
```{r tableS3, echo=TRUE, fig.height=10, fig.width=5, fig.align='center', message=FALSE, warning=TRUE}

get_table_FC_stats <- function(fc_summary_object, filter_exp_group, filter_lipid_class){
  
  tbl <- fc_summary_object$FC %>% filter(TreatmentGroup==filter_exp_group, str_detect(Compound, filter_lipid_class)) %>% 
            ungroup() %>% 
                  dplyr::select(Compound, TreatmentGroup,AnimalID,log2FC) %>% 
                  spread(key = AnimalID, value = log2FC) %>% 
                  left_join(fc_summary_object$summary %>% 
                              filter(TreatmentGroup==filter_exp_group, str_detect(Compound, filter_lipid_class)),
                            by = c("Compound","TreatmentGroup")) %>% 
            select(-lipidClass,-lipidClassBase) 
  
}

d_FC_report_Pred <- get_table_FC_stats(fc_summary_all, "PRED", "") 
d_FC_report_ACTH <- get_table_FC_stats(fc_summary_all, "ACTH", "") 
d_FC_report <- d_FC_report_Pred %>% left_join(d_FC_report_ACTH, by=(c("Compound"))) 

write_csv(d_FC_report, path =  here("output/table_S3_summary_FC_Pvalues_allsamples.csv"))

```

## 10) Supplementary Tables S5/S6/S7: Excluding specific dogs 
```{r tableS567, echo=TRUE, fig.height=10, fig.width=5, fig.align='center', message=FALSE, warning=TRUE}


fc_summary_noT4 <- get_FC_stat_summary(d_conc %>% filter(AnimalID !="T4"))
fc_summary_noT6 <- get_FC_stat_summary(d_conc %>% filter(AnimalID !="T6"))
fc_summary_noP3 <- get_FC_stat_summary(d_conc %>% filter(AnimalID !="P3"))

t_summary_noT4 <- get_table_FC_stats(fc_summary_noT4,"ACTH", "Cer")
t_summary_noT6 <- get_table_FC_stats(fc_summary_noT6,"ACTH", "PE [0-9]+")
t_summary_noP3 <- get_table_FC_stats(fc_summary_noP3,"PRED", "^DG|^TG")

write_csv(t_summary_noT4, path = here("output/table_S5_summary_noT4_Cer.csv"))
write_csv(t_summary_noT6, path = here("output/table_S6_summary_noT6_LPE-PE.csv"))
write_csv(t_summary_noP3, path = here("output/table_S7_summary_noP3_DG-TG.csv"))

knitr::kable(t_summary_noT4 %>% select(-T1:-T6) ,digits = 3, caption = "no T4")
knitr::kable(t_summary_noT6 %>% select(-T1:-T5) ,digits = 3, caption = "no T6")
knitr::kable(t_summary_noP3 %>% select(-P1:-P8),digits = 3, caption = "no P3")


```

## 11) Supplementary Figures S4 a/b/d: Average FC and p values with/without specific individuals
Comparison of fold changes and p values for all individuals and when specific
dogs were excluded from the analysis.
```{r FigS4abd, echo=TRUE, fig.height=7.8, fig.width=4.4, message=FALSE, warning=TRUE, paged.print=FALSE}
pdf(file = here("output/figure_S4abd_heatmaps.pdf"), width = 5 , height = 9)
  plot_FC_heatmap_figure(get_FC_stat_summary(d_conc %>% filter(AnimalID !="T4")), c("^Cer "))
  plot_FC_heatmap_figure(get_FC_stat_summary(d_conc %>% filter(AnimalID !="T6")), c("^LPE ","^PE [0-9]+"))
  plot_FC_heatmap_figure(get_FC_stat_summary(d_conc %>% filter(AnimalID !="P3")), c("^DG ","^TG "))
dev.off()
```

## 12) Supplementary Figure S4c: Effect of removing T6 on total levels of phospholipid species with ≤3 and ≥4 double bonds  
P values from t-tests comparing total levels of PUFA and non-PUFA containg species before and after treatments are shown after the figure.
```{r FigS4c, echo=TRUE, fig.height=5, fig.width=8, message=FALSE, warning=TRUE}
plot_nonPUFA_vs_PUFA(d_conc %>% filter(AnimalID != "T6"), "figure_S4c_nonPUFAvsPUFA.pdf")
```

## Source code of "ms_data_preprocessing.R" used in this R notebook
```{r preprocessMS_Rcode, echo=TRUE, message=FALSE, warning=TRUE, cache=TRUE, code = readLines("ms_data_preprocessing.R")}
```

## Source code of "misc_functions.R" used in this R notebook
```{r miscfunctions_Rcode, echo=TRUE, message=FALSE, warning=TRUE, cache=TRUE, code = readLines("misc_functions.R")}
```

## SessionInfo at time of knitting this R notebook
```{r sessioninfo, echo=TRUE, message=FALSE, warning=TRUE, cache=FALSE}
sessionInfo()
```
